---
title: "co2_1"
output: html_document
date: "2024-02-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Co2 emissions etc
Note: you can't just include variables from the global environment, the document has to be reproducible by itself, therefore you have to load all the data, create pltos etc here - but - it doesn't have to be visible, you can hide the code with {r, echo=FALSE}

```{r, echo=FALSE, results=FALSE, message=FALSE, warning=FALSE}
# load the libraries, data
library(tidyverse, warn.conflicts=F, quietly=T)
library(readr, warn.conflicts=F, quietly=T)
library(stringr, warn.conflicts=F, quietly=T)
library(ggExtra, warn.conflicts=F, quietly=T)
library(readxl, warn.conflicts=F, quietly=T)

co2_data <- read_csv("data/co2_data_selected_columns.csv")
co2_data_countries <- read_csv("data/co2_data_selected_columns_countries_only.csv")
```
### General analysis over time
First, we want to check if any very general trends of co2 emissions can be seen in the whole dataset. For that, we can plot the amount of co2 released throughout time, for every country in the dataset.
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
co2_xy <- ggplot(
  co2_data,
  aes(x = year, y = co2)
) +
  geom_point() +
  ggtitle("Total co2 emissions over time for each country")

co2_xy_hist <- ggMarginal(co2_xy, type="density")

co2_xy_hist
```
Clearly, we can see, since about the start of the 20th century, some countries start to overshadow others in terms of emissions. From the distribution of year values above the graph,  we can also see, for the earlier years of the dataset, a lot of values are missing.

It is also worth noting, in the dataset, beside countries present are grouped statristics for continents and important regions, which, for further analysis, were separated from the main dataset, to be analyzed separately, and possibly in comparison to individual countries.

To limit our focus to the key players, let's look at the top 10 countries, by the sum of emissions released over time.
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
co2_data_by_country <- co2_data_countries |>
  group_by(country) |>
  summarize(total_co2 = sum(co2, na.rm = TRUE))

top10_co2_countries <- co2_data_by_country |>
  arrange(desc(total_co2)) |>
  head(10) |>
  select('country')

top10_co2_countries <- unlist(top10_co2_countries, use.names=FALSE)

co2_data |>
  filter(country %in% top10_co2_countries) |>
  ggplot(
    aes(x = year, y=co2, color=country)
  ) +
  geom_line(size=2) +
  ggtitle("Total co2 emissions over time for the top 10 countries")
```
Facet by countries
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
co2_data |>
  filter(country %in% top10_co2_countries) |>
  ggplot(
    aes(x = year, y = co2, color = country)
  ) +
  geom_line(size = 2) +
  ggtitle("Co2 emissions faceted by country") +
  facet_wrap(~ country)
```
Let's look at some Individual countries

United States, seems to have had an interesting relation with co2 emissions. Very Whilst the general trend is upward, for about 15 yers now, it has turned downwards. The influence of important global events can also be seen.
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
co2_data |>
  filter(country == 'United States') |>
  ggplot(
    aes(x = year, y = co2, color = country)
  ) +
  geom_line(size = 2) +
  geom_vline(xintercept = 1930, linetype='dashed',
             color = 'blue', size = 1.2) +
  geom_text(x=1930, y = 6000, label = "The Great Depression",
            color = 'blue') +
  geom_vline(xintercept = 2008, linetype='dashed',
             color = 'green', size = 1.2) +
  geom_text(x=2000, y = 2000, label = "The 2008 Crash",
            color = 'green') +
  ggtitle("United States")
```
Germany, while on a much smaller (about 1/6) scale, follows a very similar trend of strong emission rise, then a downwards trend in the 21st century.
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
co2_data |>
  filter(country == 'Germany') |>
  ggplot(
    aes(x = year, y = co2, color = country)
  ) +
  geom_line(size = 2) +
  geom_vline(xintercept = 1933, linetype='dashed',
             color = 'brown', size = 1.2) +
  geom_text(x=1920, y = 950, label = "Hitler assigned as Chancellor",
            color = 'brown') +
  geom_vline(xintercept = 1945, linetype='dashed',
             color = 'blue', size = 1.2) +
  geom_text(x=1945, y = 900, label = "End of WW2",
            color = 'blue') +
  geom_vline(xintercept = 1989, linetype='dashed',
             color = 'green', size = 1.2) +
  geom_text(x=1989, y = 300, label = "The Fall of The Berlin Wall",
            color = 'green') +
  ggtitle("Germany")
```
For a different kind of trend, we can, for example look at China:
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
co2_data |>
  filter(country == 'China') |>
  ggplot(
    aes(x = year, y = co2, color = country)
  ) +
  geom_line(size = 2) +
  geom_vline(xintercept = 2012, linetype='dashed',
             color = 'green', size = 1.2) +
  geom_text(x=2012, y = 3000, label = "Xi Jinping",
            color = 'green') +
  ggtitle("China")
```
Bottom 10 countries, notice that for a lot of these data is missing, which might contribute to their lower scores, but the scale of emissions is so small, that's probably not the case.
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
bot10_co2_countries <- co2_data_by_country |>
  filter(total_co2 > 0) |>
  arrange(total_co2) |>
  head(10) |>
  select('country')

bot10_co2_countries <- unlist(bot10_co2_countries, use.names=FALSE)

bot10_co2_countries

saveRDS(bot10_co2_countries, file="bot10_co2_countries_names.rds")

co2_data |>
  filter(country %in% bot10_co2_countries) |>
  ggplot(
    aes(x = year, y=co2, color=country)
  ) +
  geom_line(size=2) +
  ggtitle("Total co2 emissions over time for bottom 10 countries")
```
Top 10 countries by population:
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
top10_pop_countries <- readRDS(file="top10_pop_countries_names.rds")

co2_data |>
  filter(country %in% top10_pop_countries) |>
  ggplot(
    aes(x = year, y=population, color=country)
  ) +
  geom_line(size=2) +
  ggtitle("Total population over time for top 10 countries")
```
Population and co2 emissions for United States
credits to user Megatron's [answer](https://stackoverflow.com/a/51844068)
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
max_co2_US <- co2_data |>
  filter(country == 'United States') |>
  select('co2') |>
  max(na.rm = TRUE)

max_pop_US <- co2_data |>
  filter(country == 'United States') |>
  select('population') |>
  max(na.rm = TRUE)

scaleFactor <- max_co2_US / max_pop_US

co2_data |>
  filter(country == 'United States') |>
  ggplot(
    aes(x = year)
  ) +
  geom_line(aes(y = co2, color = "co2")) +
  geom_line(aes(y = population * scaleFactor, color = "population")) +
  scale_y_continuous(
    name = "co2",
    sec.axis = sec_axis(~./scaleFactor, name = 'population')
  ) +
  ggtitle("Co2 and population for United States")
```
Top 10 countries by co2 vs rest of world
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
top10_co2_countries_data <- co2_data |>
  filter(country %in% top10_co2_countries)

top10_co2_data_by_year <- top10_co2_countries_data |>
  group_by(year) |>
  summarize(total_co2 = sum(co2, na.rm = TRUE))

not_top10_co2_countries_data <- co2_data |>
  filter(!country %in% top10_co2_countries)

not_top10_co2_data_by_year <- not_top10_co2_countries_data |>
  group_by(year) |>
  summarize(total_co2 = sum(co2, na.rm = TRUE))

top10_co2_data_by_year

ggplot(NULL, aes(x=year, y=total_co2)) +
  geom_line(data = top10_co2_data_by_year,
            aes(color = 'top 10')) +
  geom_line(data = not_top10_co2_data_by_year,
            aes(color = 'rest of world')) +
  ggtitle("Top10 vs not top 10 sum")
```
In pie form
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
top10_co2_sum <- sum(top10_co2_data_by_year$total_co2)
not_top10_co2_sum <- sum(not_top10_co2_data_by_year$total_co2)

donut_data <- data.frame(
  group = c('top 10', 'rest of world'),
  value = c(top10_co2_sum, not_top10_co2_sum)
)

ggplot(donut_data, aes(x="", y=value, fill=group)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0)
```
Co2 vs population for the top 10 countries
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
top10_co2_countries_data |>
  ggplot(
    aes(x = population, y = co2, color=country)
  ) +
  geom_point()
```
Zoomed in
A trend tends to become visible, where the slowdown of population growth corresponds to the lowering of co2 emissions. Is this a direct correlation of population decline or a sign of general 21st century societal changes in approach to co2 emissions and way of living?
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
top10_co2_countries_data |>
  ggplot(
    aes(x = population, y = co2, color=country)
  ) +
  geom_point() +
  coord_cartesian(xlim = c(0, 350000000))
```
### 2021 analysis
As 2021 is the latest year available in the dataset.
For example, checking the relation between co2 emissions and population
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
co2_pop_2021 <- co2_data_countries |>
  filter(year == 2021) |>
  ggplot(
    aes(x = population, y = co2)
  ) +
  geom_point() +
  geom_point(data = co2_data_countries
             |> filter(country %in% c('India', 'United States', 'China'))
             |> filter(year == 2021),
             pch = 21,
             size = 5,
             color = 'red') +
  geom_text(x=1.3e+09, y=1.1e+04, label='China') +
  geom_text(x=1.3e+09, y=2e+03, label='India') +
  geom_text(x=3e+08, y=4e+03, label='United States')
co2_pop_2021_dens <-ggMarginal(co2_pop_2021, type='histogram')
co2_pop_2021_dens
```
Zooming in for further insight about most of the world
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
co2_pop_2021 <- co2_pop_2021 +
  coord_cartesian(xlim = c(0, 3e+08),
                  ylim = c(0, 2e+03)) +
  geom_smooth(method = lm, level = 0.99)

co2_pop_2021_dens <-ggMarginal(co2_pop_2021, type='histogram')
co2_pop_2021_dens
```
To compare co2 emissions vs gdp, a step back to 2018 was necessary, as the dataset is missing gdp data past that year, for any country. Once again zoomed in to exclude China, United States and India
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
co2_gdp_2018 <- co2_data_countries |>
  filter(year == 2018) |>
  ggplot(
    aes(x = gdp, y = co2)
  ) +
  geom_point() +
  coord_cartesian(xlim = c(0, 5e+12),
                ylim = c(0, 2e+03)) +
  geom_smooth(method = lm, level = 0.99)

co2_gdp_2018

co2_gdp_2018_dens <-ggMarginal(co2_gdp_2018, type='histogram')
co2_gdp_2018_dens
```
2018 gdp and population
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
pop_gdp_2018 <- co2_data_countries |>
  filter(year == 2018) |>
  ggplot(
    aes(x = population, y = gdp)
  ) +
  geom_point() +
  coord_cartesian(xlim = c(0, 3e+08),
                  ylim = c(0, 5e+12)) +
  geom_smooth(method = lm, level = 0.99)

pop_gdp_2018

pop_gdp_2018_dens <-ggMarginal(pop_gdp_2018, type='histogram')
pop_gdp_2018_dens
```
### Multi-country groupings
As defined / presented in the dataset, where countries were grouped by continent, whether they belong to OECD (Organisation for Economic Co-operation and Development) and by income tiers.
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
income_split <- co2_data |>
  filter(str_detect(country, '.*income.*'))
income_split |>
  ggplot(
    aes(x = year, y = co2, color = country)
  ) +
  geom_point() +
  ggtitle(label="co2 emissions by year per income group")
```
Unfortunately, a metric for belonging to these income groups was not provided (no GDP thresholds / values provided).
Population:
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
income_split |>
  ggplot(
    aes(x = year, y = population, color = country)
  ) +
  geom_point() +
  ggtitle(label="population by year per income group")
```
Co2 vs income group boxplots
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
income_split |>
  ggplot(
    aes(x = country, y = co2)
  ) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = -90, vjust = 0.5, hjust=1)) +
  ggtitle(label="co2 distribution per income group")
```
### continents and important regions
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}

continents = c('Africa', 'Asia', 'Asia (excl. China and India)',
               'Central America (GCP)', 'Europe', 'International transport',
               'Middle East (GCP)', 'North America',
               'North America (excl. USA)', 'Oceania',
               'South America (GCP)')

co2_data_non_countries <- read_csv("data/co2_data_selected_columns_non_countries_only.csv")

co2_data_continents <- co2_data_non_countries |>
  filter(country %in% continents)

co2_data_continents |>
  ggplot(
    aes(x = year, y = co2, color = country)
  ) +
  geom_point() +
  ggtitle(label="co2 emissions per continent / important region")
```
### Per capita and participation proportions
Looking at the plot of co2 emissions per capita by year, various outliers seem to appear, muddying the whole image
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
co2_data_countries |>
  ggplot(
    aes(x = year, y = co2_per_capita)
  ) +
  geom_point() +
  ggtitle(label="Co2 per capita")
```
Looking at the plot of co2 per capita and population, the answer for these outliers seems to be appearing; a very high or a very low population can lead to unusual per capita values
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
co2_data_countries |>
  ggplot(
    aes(x = population, y = co2_per_capita)
  ) +
  geom_point() +
  ggtitle(label="Co2 per capita against population")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
co2_data_countries |>
  ggplot(
    aes(x = year, y = co2_per_unit_energy)
  ) +
  geom_point() +
  ggtitle(label="co2 per unity energy")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
co2_data_countries |>
  ggplot(
    aes(x = year, y = energy_per_capita)
  ) +
  geom_point() +
  ggtitle(label="energy_per_capita")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
co2_data_countries |>
  ggplot(
    aes(x = year, y = temperature_change_from_co2)
  ) +
  geom_point() +
  ggtitle(label="temperature_change_from_co2")
```
#### Participation / categories
What actually contributes to the total co2 emissions around the world, where do the emissions come from?
Where categories are defined as:
* coal_co2 - Annual emissions of carbon dioxide (CO2) from coal, measured in million tonnes.
* consumption_co2 - 	
Annual consumption-based emissions of carbon dioxide (CO2), measured in million tonnes.
* oil_co2
* trade_co2 - Annual net carbon dioxide (CO2) emissions embedded in trade, measured in million tonnes.
* optionally: co2_including_luc - Annual total emissions of carbon dioxide (CO2), including land-use change, measured in million tonnes.
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
categories = c('coal_co2', 'consumption_co2', 'oil_co2', 'trade_co2')

co2_data_categories <- subset(co2_data_countries, select=categories)

co2_data_categories_pivot <- tidyr::pivot_longer(co2_data_categories, categories)

co2_data_cat_pivot_group <- co2_data_categories_pivot |>
  group_by(name) |>
  summarize(
    value_sum = sum(value, na.rm = TRUE)
  )

ggplot(co2_data_cat_pivot_group,
       aes(x="", y=value_sum, fill=name)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0) +
  ggtitle(label="coal, consumption, oil, trade co2 emissions (tonnes)")
```
Contribution by input group
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
categories_country = c('country', 'coal_co2', 'consumption_co2', 'oil_co2',
                       'trade_co2')

income_data_categories <- subset(income_split, select=categories_country)

income_data_categories_pivot <- tidyr::pivot_longer(income_data_categories, categories)

income_data_cat_pivot_group <- income_data_categories_pivot |>
  group_by(country, name) |>
  summarize(
    value_sum = sum(value, na.rm = TRUE)
  )

income_data_cat_pivot_group |>
  ggplot(
    aes(country, value_sum, fill = name)
    ) +
  geom_col(position = 'fill') +
  theme(axis.text.x = element_text(angle = -90, vjust = 0.5, hjust=1)) +
  coord_cartesian(ylim = c(0, 1)) +
  ggtitle(label = "contribution by income group")
```
Contribution by continent / major region
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}
continent_data_categories <- subset(co2_data_continents, select=categories_country)

continent_data_categories_pivot <- tidyr::pivot_longer(continent_data_categories, categories)

continent_data_cat_pivot_group <- continent_data_categories_pivot |>
  group_by(country, name) |>
  summarize(
    value_sum = sum(value, na.rm = TRUE)
  )

continent_data_cat_pivot_group |>
  ggplot(
    aes(country, value_sum, fill = name)
  ) +
  geom_col(position = 'fill') +
  theme(axis.text.x = element_text(angle = -90, vjust = 0.5, hjust=1)) +
  coord_cartesian(ylim = c(0, 1)) +
  ggtitle(label = "contribution by continent")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center'}

```